### 2009-04-07 talk/icooolps2009-dotnet/clibackend.tex
Some statments rephrased, minor corrections

At the moment of writing, three backends have been implemented: one for Intel
\emph{x86} processors, one for \emph{PowerPC} processors, and one for the
\emph{CLI Virtual Machine}.  The latter is special because instead of emitting
code for a real CPU it emits code for a virtual machine\footnote{By using the 
\lstinline{Reflection.Emit} namespace and creating \lstinline{DynamicMethod}s.}: 
before being
executed, the generated code will be compiled again by .NET's own JIT
compiler.

Thus, when using the CLI backend, we actually have two layers of
JIT-compilation, each layer removing away different kinds of overhead.  By
operating at a higher level, our JIT can potentially do a better job than the
.NET one in some contexts, as our benchmarks demonstrate (see
Section~\ref{sec:benchmarks}).  On the other hand, the lower-level .NET JIT is
very good at producing machine code, much more than PyPy's own \emph{x86}
backend for example.  By combining the strengths of both we can get highly
efficient machine code.

As usual, the drawback is that programs that runs for a very short period of
time could run slower with JIT than without, due to the time spent doing the
initial (double) compilation.  It is important to underline that so far we
mostly focused making the JIT able to produce very fast code, without trying
to optimze the compilation phase itself.

@@

For a large part, implementing the CLI backend is easy and straightforward, as
there is a close correspondence between most of the operations used by
frontend's flowgraphs and the CLI instructions.  Thus, we will not dig into
details for this part.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[height=5cm]{flexswitch1}
\includegraphics[height=5cm]{flexswitch2}
\caption{An example of a flexswitch evolution: in the picture on the
  right a new block has been dynamically added.}\label{flexswitch-fig}
\end{center}
\end{figure}

@@

Unfortunately, the CLI VM does not allow modification of code which
has been already loaded and linked, therefore the simplest approach
taken for low level architectures does not work.

@@

\item Each either primary or secondary method implements a certain
  number of blocks, all belonging to the same flow graph.
\end{itemize}

@@

A link is called \emph{internal} if it connects two blocks contained
in the same method,
\emph{external} otherwise.

@@

The code\footnote{For simplicity we write C\# code instead of
the actual IL bytecode.} generated for the dispatcher of methods
is similar to the following fragment: 
\begin{small}
\begin{lstlisting}[language={[Sharp]C}]
// dispatch block
int methodid = (blockid && 0xFFFF0000) >> 16; 
int blocknum = blockid && 0x0000FFFF;         
if (methodid != MY_METHOD_ID) {
  // jump_to_ext 
  ...
}
switch(blocknum) {
  case 0: goto block0;
  case 1: goto block1;
  default: throw new Exception("Invalid block id");
}
\end{lstlisting}
\end{small}
If the next block to be executed is implemented in the same method
({\small\lstinline{methodid == MY_METHOD_ID}}), then the appropriate
jump to the corresponding code is executed.
Otherwise, the \lstinline{jump_to_ext}
part of the dispatcher has to be executed, which is implemented differently 
by primary and secondary methods.

The primary method is responsible for the bookkeeping of the secondary
methods which are added to the same graph dynamically. This can be 
simply implemented with an array mapping method id of secondary methods
to the corresponding delegate. Therefore, the primary methods contain
the following \lstinline{jump_to_ext} code (where
\lstinline{FlexSwitchCase} is the type of delegates for secondary methods):
\begin{small}
\begin{lstlisting}[language={[Sharp]C}] 
// jump_to_ext
FlexSwitchCase meth = method_map[methodid];
blockid = meth(blockid, ...); // execute the method
goto dispatch_block;
\end{lstlisting}
\end{small}
Each secondary method returns the block id of the next block to be
executed; therefore, after the secondary method has returned, the
dispatcher of the primary method will be executed again to jump
to the correct next block.

@@

Therefore, the solution we came up with is defining a class
\lstinline{InputArgs} for passing sequences of arguments whose length
and type is variable.
\begin{small}
\begin{lstlisting}[language={[Sharp]C}] 
public class InputArgs {
  public int[] ints;
  public float[] floats;
  public object[] objs;
  ...
}
\end{lstlisting}
\end{small}
Unfortunately, with this solution passing arguments to external links
becomes quite slow:
\begin{itemize}
\item When writing arguments, array re-allocation may be needed in
  case the number of arguments exceeds the dimension of the
  array. Furthermore the VM will always perform bound-checks, even
  when the size is explicitly checked in advance;

@@

\subsubsection{Implementation of flexswitches}
Finally, we can have a look at the implementation of flexswitches.
The following snippet shows the special case of integer flexswitches.
\begin{small}
\begin{lstlisting}[language={[Sharp]C}] 
public class IntLowLevelFlexSwitch: 
                        BaseLowLevelFlexSwitch {
  public uint default_blockid = 0xFFFFFFFF;
  public int numcases = 0;
  public int[] values = new int[4];
  public FlexSwitchCase[] cases = 
                          new FlexSwitchCase[4];

@@

  public uint execute(int value, InputArgs args) {
    for(int i=0; i<numcases; i++)
      if (values[i] == value)
        return cases[i](0, args);
    return default_blockid;
  }
}
\end{lstlisting}
\end{small}
The mapping from integers values to delegates (pointing to secondary
methods) is just implemented by the two arrays \lstinline{values} and
\lstinline{cases}. Method \lstinline{add_case} extends the mapping
whenever a new case is added to the flexswitch.

>>>>

At the moment of writing, three backends have been implemented: one for Intel
\emph{x86} processors, one for \emph{PowerPC} processors, and one for the
\emph{CLI Virtual Machine}.  The latter is special because instead of emitting
code for a real CPU it emits code for a virtual machine\footnote{By using the 
\lstinline{Reflection.Emit} namespace and creating \lstinline{DynamicMethod}s.}: 
before being executed, the generated code will be compiled again by the .NET JIT
compiler.

Thus, when using the CLI backend, we actually have two JIT compilers at two different levels
layers, each one specialized in different kinds of optimization.
By operating at a higher level, our JIT can potentially do a better job 
in some contexts, as our benchmarks demonstrate (see
Section~\ref{sec:benchmarks}).  On the other hand, the lower-level .NET JIT is
very good at producing machine code, much more than PyPy's own \emph{x86}
backend, for example.  By combining the strengths of both we can get highly
efficient machine code.

As usual, the drawback is that programs that runs for a very short period of
time could run slower with JIT than without, due to the time spent doing the
initial (double) compilation.  
Finally, it is important to underline that while we have directed our efforts 
to generate a JIT compiler able to emit very efficient code,
the performance of the compiler itself has been neglected so far, but  
it is certainly an issue which will have to be considered in our future research.

@@

For a large part, implementing the CLI backend is easy and straightforward, as
there is a close correspondence between most of the operations used by
frontend's flowgraphs and the CLI instructions.  Thus, we will not go into
details for this part.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[height=5cm]{flexswitch1}
\includegraphics[height=5cm]{flexswitch2}
\caption{An example of a flexswitch evolution: in the picture on the
  right block 7 has been dynamically added.}\label{flexswitch-fig}
\end{center}
\end{figure}

@@

Unfortunately, CLI does not allow modification of code which
has been already loaded and linked, therefore the simplest approach
taken for low level architectures does not work.

@@

\item Each either primary or secondary method implements a certain
  number of blocks, all belonging to the same flow graph.

\item The parameters of a method correspond to the arguments of what we call the \emph{initial block} of the method.

\end{itemize}

@@

A link is called \emph{internal} if it connects two blocks implemented
by the same method,
\emph{external} otherwise.

@@

The code\footnote{For simplicity we write C\# code instead of
the actual IL bytecode.} generated for the dispatcher of methods
is similar to the following fragment: 
\begin{lstlisting}[language={[Sharp]C}]
// dispatch block
int methodid = (blockid && 0xFFFF0000) >> 16; 
int blocknum = blockid && 0x0000FFFF;         
if (methodid != MY_METHOD_ID) {
  // jump_to_ext 
  ...
}
switch(blocknum) {
  case 0: goto block0;
  case 1: goto block1;
  default: throw new Exception("Invalid block id");
}
\end{lstlisting}
If the next block to be executed is implemented in the same method
(\lstinline{methodid == MY_METHOD_ID}), then the appropriate
jump to the corresponding code is executed.
Otherwise, the \lstinline{jump_to_ext}
part of the dispatcher has to be executed, which is implemented differently 
by primary and secondary methods.

The primary method is responsible for the bookkeeping of the secondary
methods which are added to the same graph dynamically. This can be 
simply implemented with an array mapping method id of secondary methods
to the corresponding delegate. Therefore, the primary methods contain
the following \lstinline{jump_to_ext} code (where
\lstinline{FlexSwitchCase} is the type of delegates for secondary methods):
\begin{lstlisting}[language={[Sharp]C}] 
// jump_to_ext
FlexSwitchCase meth = method_map[methodid];
blockid = meth(blockid, ...); // execute the method
goto dispatch_block;
\end{lstlisting}
Each secondary method returns the block id of the next block to be
executed; therefore, after the secondary method has returned, the
dispatcher of the primary method will be executed again to jump
to the correct next block.

@@

Therefore, the solution we came up with is defining a class
\lstinline{InputArgs} for passing sequences of arguments whose length
and type is variable.
\begin{lstlisting}[language={[Sharp]C}] 
public class InputArgs {
  public int[] ints;
  public float[] floats;
  public object[] objs;
  ...
}
\end{lstlisting}
Unfortunately, with this solution passing arguments to external links
becomes quite slow:
\begin{itemize}
\item When writing arguments, array re-allocation may be needed in
  case the number of arguments exceeds the dimension of the
  array. Furthermore the VM will always perform bound-checks, even
  when the size is explicitly checked in advance;

@@

\subsubsection{Implementation of flexswitches}
Finally, we can have a look at the implementation of flexswitches.
The following snippet shows the special case of integer flexswitches.
\begin{lstlisting}[language={[Sharp]C}] 
public class IntLowLevelFlexSwitch: 
                        BaseLowLevelFlexSwitch {
  public uint default_blockid = 0xFFFFFFFF;
  public int numcases = 0;
  public int[] values = new int[4];
  public FlexSwitchCase[] cases = 
                          new FlexSwitchCase[4];

@@

  public uint execute(int value, InputArgs args) {
    for(int i=0; i<numcases; i++)
      if (values[i] == value)
        return cases[i](0, args);
    return default_blockid;
  }
}
\end{lstlisting}
The mapping from integers values to delegates (pointing to secondary
methods) is just implemented by the two arrays \lstinline{values} and
\lstinline{cases}. Method \lstinline{add_case} extends the mapping
whenever a new case is added to the flexswitch.

### 2009-04-07 talk/icooolps2009-dotnet/conclusion.tex
Some statments rephrased, minor corrections

Moreover, the idea of promotion is a generalization of \emph{Polymorphic
  Inline Caches} \cite{hoelzle_optimizing_1991}, as well as the idea of using
runtime feedback to produce more efficient code
\cite{hoelzle_type_feedback_1994}.

@@

In this paper we gave an overview of PyPy's JIT compiler generator,
which can automatically turn an interpreter into a JIT
compiler, requiring the language developers to only add few \texttt{hint}s to
guide the generation process.

Then, we presented the CLI backend for PyPy's JIT compiler generator, whose
goal is to produce .NET bytecode at runtime.  We showed how it is possible to
circumvent intrinsic limitations of the virtual machine to implement
flexswitches.  As a result, we proved that the idea of \emph{JIT layering} is
worth of further exploration, as it makes possible for dynamically typed
languages to be even faster than their statically typed counterpart in some
circumstances.

As a future work, we want to explore different strategies to make the frontend
producing less code, maintaining comparable or better performances.  In
particular, we are working on a way to automatically detect loops in the user
code, as tracing JITs do \cite{gal_hotpathvm_2006}.  By compilining whole
loops at once, the backends should be able to produce better code than today.

At the moment, some bugs and minor missing features prevent the CLI JIT
backend to handle more complex languages such as Python and Smalltalk.  We are
confident that once these problems will be fixed, we will get performance
results comparable to TLC, as the other backends already demonstrate
\cite{PyPyJIT}.  Moreover, if the current implementation of flexswitches will
prove to be too slow for some purposes, we want to explore alternative
implementation strategies, also considering the new features that might be
integrated into virtual machines.

>>>>

The idea of promotion is a generalization of \emph{Polymorphic
  Inline Caches} \cite{hoelzle_optimizing_1991}, as well as the idea of using
runtime feedback to produce more efficient code
\cite{hoelzle_type_feedback_1994}.

@@

In this paper we gave an overview of PyPy's JIT compiler generator,
which can automatically turn an interpreter into a JIT
compiler, requiring the language developers to only add few hints to
guide the generation process.

Then, we presented the CLI backend for PyPy's JIT compiler generator, whose
goal is to produce .NET bytecode at runtime.  We showed how it is possible to
circumvent intrinsic limitations of the virtual machine to implement
flexswitches.  As a result, we proved that the idea of \emph{JIT layering} is
worth of further exploration, as it makes possible for dynamically typed
languages to be even faster than their statically typed counterpart in some
cases.

As a future work, we want to explore different strategies to make the frontend
producing less code, maintaining comparable or better performances.  In
particular, we are working on a way to automatically detect loops in the user
code, as tracing JITs do \cite{gal_hotpathvm_2006}.  By compiling whole
loops at once, the backends should be able to produce better code.

At the moment, some bugs and minor missing features prevent the CLI JIT
backend to handle more complex languages such as Python and Smalltalk.  We are
confident that once these problems will be fixed, we will get performance
results comparable to TLC, as the other backends already demonstrate
\cite{PyPyJIT}.  However, if the current implementation of flexswitches will
turn out to be too slow for some purposes, alternative
implementation strategies could be explored by considering the novel features 
offered the new generation of virtual machines.
